---
title: "Objective 0: Data Preparation"
author: "Zoe Penno"
date: "2025-07-03"
output: html_document
---

# Objective 0: Data Preparation

This script prepares inputs for downstream GLM and GAM analyses using only the three raw data files:
1. data/raw_data/images_public.csv
2. data/raw_data/stations_public.csv
3. data/raw_data/tide_data.csv

Outputs created here (in data/processed_data):
- bear_count_motion.rds — site-level counts of black bear foraging days and camera effort (used by 01–04)
- detections_with_tide.rds — detection-level data joined with hourly tide height (used by 08_tide_GAM)

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(readr)
library(here)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load Data
```{r load-data}
images   <- read_csv(here("data", "raw_data", "images_public.csv"))
stations <- read_csv(here("data", "raw_data", "stations_public.csv"))
tides    <- read_csv(here("data", "raw_data", "tide_data.csv"), show_col_types = FALSE)
```

## 2. Clean Images and Join Stations
```{r clean-images}
exclude_species <- c("Homo sapiens", "Canis familiaris")

images_clean <- images %>%
  filter(!latin_name %in% exclude_species,
         !is.na(latin_name),
         latin_name != "Unknown species") %>%
  mutate(
    event_time = coalesce(ymd_hms(exif_timestamp, tz = "UTC"),
                          ymd_hms(image_timestamp, tz = "UTC")),
    event_hour = floor_date(event_time, "hour")
  ) %>%
  filter(!is.na(event_time)) %>%
  left_join(stations, by = "station_id")
```

## 3. Prepare Hourly Tide Data
```{r prep-tides}
tide_hourly <- tides %>%
  mutate(
    tide_datetime = ymd_hm(Obs_date, tz = "UTC"),
    tide_hour     = floor_date(tide_datetime, "hour")
  ) %>%
  transmute(tide_hour, tide_height = `SLEV(metres)`)
```

## 4. Attach Tide Height to Detections
```{r attach-tide}
detections_with_tide <- images_clean %>%
  left_join(tide_hourly, by = c("event_hour" = "tide_hour")) %>%
  mutate(
    julian_day        = yday(event_time),
    tide_bin          = cut(tide_height, breaks = c(-Inf, 1, 2.5, Inf),
                            labels = c("Low", "Mid", "High"), include.lowest = TRUE),
    bear_detected     = if_else(latin_name == "Ursus americanus", 1L, 0L),
    foraging_detected = if_else(latin_name == "Ursus americanus" & behaviour == "foraging", 1L, 0L, missing = 0L)
  )

write_rds(detections_with_tide, here("data", "processed_data", "detections_with_tide.rds"))
```

## 5. Create bear_count_motion for GLM scripts
```{r bear-count-motion}
# Daily presence/absence of foraging bears
bear_motion <- detections_with_tide %>%
  filter(image_trigger == "Motion Detection",
         foraging_detected == 1L) %>%
  mutate(date = as_date(event_time)) %>%
  group_by(station_id, date) %>%
  summarise(foraging_detected = 1, .groups = "drop")

# Camera effort (any image) by day
active_days <- detections_with_tide %>%
  mutate(date = as_date(event_time)) %>%
  distinct(station_id, date) %>%
  mutate(active = 1)

# Join detections to effort and restrict to May–October
bear_daily <- active_days %>%
  left_join(bear_motion, by = c("station_id", "date")) %>%
  mutate(foraging_detected = replace_na(foraging_detected, 0)) %>%
  filter(month(date) %in% 5:10)

# Aggregate to station-level counts
bear_count_motion <- bear_daily %>%
  group_by(station_id) %>%
  summarise(
    bear_days = sum(foraging_detected),
    total_days = n(),
    .groups = "drop"
  ) %>%
  left_join(stations, by = "station_id")

write_rds(bear_count_motion, here("data", "processed_data", "bear_count_motion.rds"))
```

# End of Script
