---
title: "01 Data Exploration and Cleaning"
author: "Zoe Penno"
date: "2025-06-30"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(car)
library(corrplot)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Reproducibility
set.seed(123)
```


# Introduction

This script performs exploratory data analysis and visualization for the Barkley Sound black bear camera trap study. It loads the processed `bear_count_motion.rds` dataset (created in 00_data_preparation.Rmd) and generates summary statistics and quick plots to explore site-level covariates and bear foraging patterns.

The workflow includes:

- **Summarizing** detection counts and camera effort across stations.
- **Visualizing** the distribution of key environmental variables (substrate type, development, freshwater, forest cover, salmon distance).
- **Checking collinearity** among predictors using VIF and correlation analysis.
- **Documenting camera operability** by site and over time.

# Load bear_count_motion produced in 00_data_preparation
```{r}
bear_count_motion <- read_rds(here("data", "processed_data", "bear_count_motion.rds"))

# Print summary table for supplement
summary_tbl <- bear_count_motion %>%
  summarise(
    mean_bear_days = mean(bear_days),
    sd_bear_days = sd(bear_days),
    min_bear_days = min(bear_days),
    max_bear_days = max(bear_days),
    n_stations = n()
  )
print(summary_tbl)
```


# Quick Plots of Key Variables
```{r plots, fig.width=7, fig.height=4}
# Total bear detections by station
ggplot(
  bear_count_motion,
  aes(
    x = reorder(as.factor(station_id), -bear_days),
    y = bear_days
  )
) +
  geom_col(fill = "#333333") +
  labs(
    x = "Station",
    y = "Total Bear Detections (May–Oct)",
    title = "Total Bear Detections by Station"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Distribution of substrate types
ggplot(bear_count_motion, aes(x = substrate_group, fill = substrate_group)) +
  geom_bar() +
  labs(x = "Substrate Type", y = "Count", title = "Distribution of Substrate Types") +
  theme(legend.position = "none")

# Human disturbance
ggplot(bear_count_motion, aes(x = factor(human_development))) +
  geom_bar(fill = "#b22222") +
  labs(x = "Human Disturbance (0=No, 1=Yes)", y = "Count", title = "Human Disturbance by Station")

# Distribution of freshwater presence
ggplot(bear_count_motion, aes(x = factor(freshwater))) +
  geom_bar(fill = "#1f77b4") +
  labs(x = "Freshwater Presence (0=No, 1=Yes)", y = "Count", title = "Freshwater Presence by Station") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Coniferous forest cover per site
ggplot(bear_count_motion, aes(x = reorder(as.factor(station_id), prop_coastal_coniferous_forest), y = prop_coastal_coniferous_forest)) +
    geom_col(fill = "#228B22") +
    labs(
        x = "Station",
        y = "Proportion Coniferous Forest",
        title = "Coniferous Forest Cover by Station"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Distance to salmon streams
ggplot(bear_count_motion, aes(x = reorder(as.factor(station_id), dist_to_salmon_km), y = dist_to_salmon_km)) +
    geom_col(fill = "#1f77b4") +
    labs(
        x = "Station",
        y = "Distance to Salmon Streams (m)",
        title = "Distance to Salmon Streams by Station"
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Assumption checks before modeling: collinearity and distribution checks
```{r collinearity-checks}
# Check for collinearity among predictors using Variance Inflation Factor (VIF)
bear_count_motion$substrate_group <- as.factor(bear_count_motion$substrate_group)

table(bear_count_motion$substrate_group)

# Choose only numeric and factor predictors
vif_data <- bear_count_motion %>%
  select(
    bear_days,
    dist_to_salmon_km,
    prop_coastal_coniferous_forest,
    prop_coastal_shrubland,
    prop_coastal_meadow_grassland,
    freshwater,
    human_development,
    substrate_group
  ) %>%
  drop_na()

# Fit a linear model (VIF requires lm(), not glm)
vif_model <- lm(bear_days ~ ., data = vif_data)

# Compute VIF
vif_results <- car::vif(vif_model)
print(vif_results)

# Correlation matrix for numeric predictors
cont_vars <- bear_count_motion %>%
  select(
    dist_to_salmon_km,
    starts_with("prop_")
  ) %>%
  drop_na()

cor_matrix <- cor(cont_vars)
round(cor_matrix, 2)

# Visualize correlation matrix
corrplot(cor(cont_vars, use = "complete.obs"), method = "color", addCoef.col = "black")
```
Based on the VIF results and correlation matrix, we can see that the `prop_coastal_coniferous_forest` and `prop_coastal_shrubland` variables exhibit multicollinearity. Only `prop_coastal_coniferous_forest` and `prop_coastal_meadow_grassland` will be used in the model.

# Camera Operability Information
```{r camera-operability}
# Load raw images data to assess camera operability
images <- read_csv(here("data", "raw_data", "images_public.csv")) 

# Add date column for effort tracking
images_with_date <- images %>%
  mutate(date = as_date(exif_timestamp))

# Per-station operability summary (unique active days)
camera_summary <- images_with_date %>%
  distinct(station_id, date) %>%
  group_by(station_id) %>%
  summarise(
    installation_date = min(date, na.rm = TRUE),
    removal_date      = max(date, na.rm = TRUE),
    total_camera_days = n(),
    .groups = "drop"
  ) %>%
  arrange(station_id)

# Summary statistics
total_effort_days <- sum(camera_summary$total_camera_days)
mean_effort_days  <- mean(camera_summary$total_camera_days)
range_effort_days <- range(camera_summary$total_camera_days)

cat(sprintf(
  "Total camera-days of effort: %d\nMean (per site): %.1f  |  Range: %d–%d\nNumber of sites: %d\n",
  total_effort_days, mean_effort_days, range_effort_days[1], range_effort_days[2], nrow(camera_summary)
))

# Effort by month
effort_by_month <- images_with_date %>%
  distinct(station_id, date) %>%
  mutate(month = floor_date(date, "month")) %>%
  count(month, name = "camera_days") %>%
  arrange(month)

# Per-station-per-month effort
effort_site_month <- images_with_date %>%
  distinct(station_id, date) %>%
  mutate(month = floor_date(date, "month")) %>%
  count(station_id, month, name = "camera_days") %>%
  arrange(station_id, month)

# Save effort summaries
write_csv(camera_summary, here("data", "processed_data", "camera_operability_by_site.csv"))
write_csv(effort_by_month, here("data", "processed_data", "effort_by_month.csv"))
write_csv(effort_site_month, here("data", "processed_data", "effort_by_site_month.csv"))
```

# Session Info
```{r session-info, echo=FALSE}
sessionInfo()
```
