---
title: "02 GLM Model Selection and Inference"
author: "Zoe Penno"
date: "2025-06-30"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(broom.mixed)
library(AICcmodavg)
library(here)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Reproducibility
set.seed(123)
```

# Introduction

This script fits generalized linear models (GLMs) to identify predictors of black bear foraging use of intertidal habitat, using `bear_days` as the response variable. Candidate predictors are based on ecological rationale and site-level covariates.

The workflow includes:
- **Loading** in the data created in 01_data_exploration (bear_count_motion.rds).
- **Fitting** a set of negative binomial GLMMs which represent competing hypotheses.
- **Comparing** the competing hypotheses to determine the best fitting model.
- **Summarizing** the top model and visualizing parameter estimates.
- **Model diagnostics** to check assumptions and fit.
- **Cross-validation** to assess predictive performance.

The top model forms the foundation of the spatial inference presented in my thesis.

# Data Import
```{r data-import}
bear_count_motion <- read_rds(here("data","processed_data", "bear_count_motion.rds"))
```

# Overdispersion Check

Initial data exploration revealed overdispersion in a Poisson model, so we use negative binomial (nbinom2) family for all candidate models.

All Candidate models will use bear_count_motion dataset. All models will have bear_days (count) as the response, total_days as an offset and family = negative binomial.

Competing hypotheses:

H1:Multifactor hypothesis (global model)
Bear use is shaped by a combination of local resource availability and landscape-scale features, including substrate type, freshwater access, forest cover, and development.

H2: Resource availability hypothesis
Bear use is driven primarily by foraging-related resources such as substrate type, freshwater access, and proximity to salmon-bearing streams.

H3: Landscape disturbance hypothesis
Bear use is primarily influenced by landcover features that may facilitate or limit access to intertidal habitat, such as coniferous forest cover and human development.

H4:Site-level habitat features hypothesis
Bear use is shaped by physical characteristics of the site itself, including substrate and adjacent landcover (development and forest), without considering broader resource features like salmon.

H5: Null Hypothesis
Bear use is random with respect to environmental and landscape variables.

```{r model-selection}
# H1: Global model (Multifactor hypothesis)
model_global <- glmmTMB(
  bear_days ~ substrate_group + freshwater + prop_coastal_coniferous_forest + human_development + dist_to_salmon_km +
    offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)

# H2: Resource availability hypothesis
model_resource <- glmmTMB(
  bear_days ~ substrate_group + freshwater + dist_to_salmon_km +
    offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)

# H3: Landscape disturbance hypothesis
model_disturbance <- glmmTMB(
  bear_days ~ prop_coastal_coniferous_forest + human_development +
    offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)

# H4: Site-level habitat features hypothesis
model_site_level <- glmmTMB(
  bear_days ~ substrate_group + prop_coastal_coniferous_forest + human_development +
    offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)

# H5: Null hypothesis
model_null <- glmmTMB(
  bear_days ~ 1 + offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)

# Post-hoc model analyzing the impact of shrubland (not included in main hypotheses)
model_shrubland <- glmmTMB(
  bear_days ~ prop_coastal_shrubland + human_development +
    offset(log(total_days)),
  family = nbinom2(),
  data = bear_count_motion
)
```

# Compare the models
```{r model-comparison}
model_set <- list(
  H1_global = model_global,
  H2_resource = model_resource,
  H3_disturbance = model_disturbance,
  H4_site_level = model_site_level,
  H5_null = model_null,
  H6_shrubland = model_shrubland
)

# Print full AIC table for transparency
aic_table <- aictab(model_set)
print(aic_table)

# Summarize the best model
if (length(model_set) > 0) {
  top_models <- model_set[which.min(sapply(model_set, AIC))]
  print(summary(top_models[[1]]))
}

# Print tidy effect size table for supplement
if (exists("top_models")) {
  broom.mixed::tidy(top_models[[1]], conf.int = TRUE) %>%
    filter(effect == "fixed") %>%
    select(term, estimate, std.error, conf.low, conf.high, p.value) %>%
    print()
}
```

# Save best model object for reproducibility
```{r save-model}
# Save the best model for later use
best_model <- top_models[[1]]

saveRDS(best_model, here("data", "processed_data", "glmm_best_model.rds"))
```


# Session Info
```{r session-info, echo=FALSE}
sessionInfo()
```
