---
title: "04 GLMM Figures"
author: "Zoe Penno"
date: "2025-07-02"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(glmmTMB)
library(broom.mixed)
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Reproducibility
set.seed(123)
```

# Introduction

This script visualizes the results of the GLMM analysis.

# Load data and best model
```{r load-model}
best_model <- readRDS(here("data", "processed_data", "glmm_best_model.rds"))
bear_count_motion <- read_rds(here("data", "processed_data", "bear_count_motion.rds"))
```

# Visualize Parameter Estimates of the Best Model
```{r plot-params, fig.width=7, fig.height=5}

# Plot effect sizes for the best model with updated aesthetics
plot_df <- broom.mixed::tidy(best_model, conf.int = TRUE) %>%
  filter(effect == "fixed") %>%
  mutate(
    term = case_match(term,
      "freshwater" ~ "Freshwater Present",
      "prop_coastal_coniferous_forest" ~ "Coniferous Forest Cover",
      "human_development" ~ "Human Development",
      "dist_to_salmon_km" ~ "Distance to Salmon (km)",
      .default = term
    ),
    term = forcats::fct_reorder(term, estimate)
  )

p <- ggplot(plot_df, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(size = 1.2, color = "#0072B2") +
  geom_point(size = 1.8, color = "#0072B2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  coord_flip() +
  labs(
    x = "",
    y = "Estimate (log scale)"
  ) +
  scale_x_discrete() +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14, face = "plain"),
    axis.title.y = element_text(size = 14, face = "plain"),
    legend.position = "none",
    strip.text = element_text(size = 13, face = "plain"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.margin = margin(10, 10, 10, 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text = element_text(color = "black"),
    axis.text.y = element_text(margin = margin(r = 10))
  )

print(p)

# Save the plot
ggsave(
  here("figures", "glmm_parameter_estimates.png"),
  plot = p,
  width = 8,
  height = 6,
  dpi = 300
)

```

# Visualize Odds Ratio of the Best Model
```{r}
# Visualize Odds Ratios (exponentiated GLM coefficients)
# NOTE: For binomial-logit models these are odds ratios;
# for negative binomial models these are incidence rate ratios.

or_df <- plot_df %>%
  mutate(
    OR = exp(estimate),
    OR_low = exp(conf.low),
    OR_high = exp(conf.high)
  )

odds_plot <- ggplot(or_df, aes(x = term, y = OR)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_point(size = 3, color = "#0072B2") +
  geom_linerange(aes(ymin = OR_low, ymax = OR_high), linewidth = 0.6, color = "#0072B2") +
  scale_x_discrete(limits = rev) +
  coord_flip() +
  ylab("Odds ratio") +
  theme_classic(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)  # <-- adds box
  )

print(odds_plot)


# Save the plot
ggsave(
  here("figures", "glmm_odds_ratios.png"),
  plot = odds_plot,
  width = 7,
  height = 5,
  dpi = 300
)

```

# Predicted Probability vs Proportion of Coniferous Forest Cover
```{r predicted-probability, fig.width=7, fig.height=5}

# Create new data for prediction across the observed range of coniferous forest cover
model_vars <- all.vars(formula(best_model))
model_vars <- setdiff(model_vars, c("bear_days", "prop_coastal_coniferous_forest", "log(total_days)"))

newdata <- tibble(
  prop_coastal_coniferous_forest = seq(0, 1, by = 0.01)
)

# Add other predictors at median (numeric) or reference (factor) values
for (col in model_vars) {
  if (is.numeric(bear_count_motion[[col]])) {
    newdata[[col]] <- median(bear_count_motion[[col]], na.rm = TRUE)
  } else if (is.factor(bear_count_motion[[col]])) {
    newdata[[col]] <- levels(bear_count_motion[[col]])[1]
  }
}
newdata$total_days <- median(bear_count_motion$total_days, na.rm = TRUE)

# Predict (on link scale), then convert to response
preds <- predict(
  best_model,
  newdata = newdata,
  type = "link",
  se.fit = TRUE,
  re.form = NA
)

newdata <- newdata %>%
  mutate(
    fit = exp(preds$fit),
    lower = exp(preds$fit - 1.96 * preds$se.fit),
    upper = exp(preds$fit + 1.96 * preds$se.fit)
  )


# Create the plot
p <- ggplot(newdata, aes(x = prop_coastal_coniferous_forest, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = "#0072B2") +
  geom_line(color = "#0072B2", size = 1.5) +
  geom_rug(data = bear_count_motion,
         aes(x = prop_coastal_coniferous_forest),
         inherit.aes = FALSE,
         sides = "b",          # bottom only
         length = unit(0.02, "npc"),
         alpha = 0.6,
         colour = "black") +
  labs(
    x = "Proportion Coastal Coniferous Forest",
    y = "Predicted Probability of Bear Foraging (≥ 1 day)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 14, face = "plain"),
    axis.title.y = element_text(size = 14, face = "plain"),
    axis.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.margin = margin(10, 30, 10, 10),
    panel.grid = element_blank()
  )

print(p)

# Save the plot
ggsave(
  here("figures", "predicted_prob_vs_coniferous_forest.png"),
  plot = p,
  width = 7,
  height = 5,
  dpi = 300
)

```

# Predicted Probability vs Human Development
```{r predicted-probability-human-development, fig.width=7, fig.height=5}
# Create new data for prediction across human_development = 0 and 1
model_vars <- all.vars(formula(best_model))
model_vars <- setdiff(model_vars, c("bear_days", "human_development", "log(total_days)"))

newdata <- tibble(
  human_development = c(0, 1)
)

# Add other predictors at median (numeric) or reference (factor) values
for (col in model_vars) {
  if (is.numeric(bear_count_motion[[col]])) {
    newdata[[col]] <- median(bear_count_motion[[col]], na.rm = TRUE)
  } else if (is.factor(bear_count_motion[[col]])) {
    newdata[[col]] <- levels(bear_count_motion[[col]])[1]
  }
}
newdata$total_days <- median(bear_count_motion$total_days, na.rm = TRUE)

# Predict on the link scale and convert to response
preds <- predict(
  best_model,
  newdata = newdata,
  type = "link",
  se.fit = TRUE,
  re.form = NA
)

newdata <- newdata %>%
  mutate(
    fit = exp(preds$fit),
    lower = exp(preds$fit - 1.96 * preds$se.fit),
    upper = exp(preds$fit + 1.96 * preds$se.fit)
  )

# Plot
p <- ggplot(newdata, aes(x = factor(human_development), y = fit)) +
  geom_point(size = 3, color = 'black') +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = 'black') +
  labs(
    x = "Human Development (0 = Absent, 1 = Present)",
    y = "Predicted Probability of Bear Foraging (≥ 1 day)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 14, face = "plain"),
    axis.title.y = element_text(size = 14, face = "plain"),
    axis.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.margin = margin(10, 30, 10, 10),
    panel.grid = element_blank()
  )

print(p)

# Save the plot
ggsave(
  here("figures", "predicted_prob_vs_human_development.png"),
  plot = p,
  width = 7,
  height = 5,
  dpi = 300
)

```

# Camera Trap Operability Table
```{r camera-trap-operability, echo=FALSE}
# Load camera operability summary from processed_data (created in 01_data_exploration)
camera_summary <- read_csv(here("data", "processed_data", "camera_operability_by_site.csv"))

# Show the summary table
knitr::kable(camera_summary, caption = "Camera Trap Operability by Site")
```

# Camera Trap Operability Gantt Chart
```{r camera-trap-gantt, fig.width=10, fig.height=8}
# Load images data for activity timeline
images <- read_csv(here("data", "raw_data", "images_public.csv"))

# Show camera activity periods (with gaps) for each site
activity_df <- images %>%
  mutate(date = as_date(exif_timestamp)) %>%
  distinct(station_id, date) %>%
  arrange(station_id, date) %>%
  group_by(station_id) %>%
  mutate(
    gap = as.integer(date - lag(date, default = first(date) - 1)),
    run_id = cumsum(if_else(is.na(gap) | gap > 1, 1, 0))
  ) %>%
  group_by(station_id, run_id) %>%
  summarise(
    start = min(date),
    end = max(date),
    .groups = "drop"
  )

ggplot(activity_df, aes(y = reorder(station_id, start), x = start, xend = end)) +
  geom_segment(aes(xend = end, yend = station_id), size = 6, color = "#0072B2") +
  labs(
    x = "Date",
    y = "Station"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 14, face = "plain"),
    axis.title.y = element_text(size = 14, face = "plain"),
    axis.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    plot.margin = margin(10, 10, 10, 10)
  )

# Save the Gantt chart
ggsave(
  here("figures", "camera_trap_gantt_chart.png"),
  width = 10,
  height = 8,
  dpi = 300
)
```

# Session Info
```{r session-info, echo=FALSE}
sessionInfo()
```
