---
title: "05 GAM Models: Seasonal and Tide"
author: "Zoe Penno"
date: "2025-06-30"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(mgcv)
library(gratia)
library(here)
library(patchwork)
library(broom)
library(glmmTMB)
library(performance)
library(MuMIn)
library(yardstick)
library(DHARMa)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Reproducibility
set.seed(123)
```

# Overview

This combined script fits two GAM analyses:
- **Seasonal GAM (foraging only)** using raw images to show temporal trends across stations.
- **Tide GAM (bear presence)** using detections_with_tide.rds to assess tide and seasonal effects.

# Part A: Seasonal GAM (foraging only)

## Data prep
```{r seasonal-data}
images <- read_csv(here("data", "raw_data", "images_public.csv"))

# Active camera days during peak season
active_days <- images %>%
  mutate(date = as_date(exif_timestamp)) %>%
  filter(month(date) %in% 5:10) %>%
  distinct(station_id, date) %>%
  mutate(station_id = as.factor(station_id))

# Days with foraging detections
foraging_days <- images %>%
  filter(image_trigger == "Motion Detection",
         latin_name == "Ursus americanus",
         behaviour == "foraging") %>%
  mutate(date = as_date(exif_timestamp)) %>%
  distinct(station_id, date)

# Binary response per day
daily_df <- active_days %>%
  left_join(foraging_days %>% mutate(foraging_flag = 1), by = c("station_id", "date")) %>%
  mutate(
    foraging_detected = if_else(is.na(foraging_flag), 0, 1),
    julian_day = yday(date),
    station_id = as.factor(station_id)
  ) %>%
  select(station_id, date, julian_day, foraging_detected)
```

## Fit seasonal GAM
```{r seasonal-gam}
gam_season <- gam(
  foraging_detected ~ s(julian_day, by = station_id) + s(station_id, bs = "re"),
  data = daily_df,
  family = binomial
)
summary(gam_season)
```

## Plot significant smooths
```{r seasonal-plots, fig.width=7, fig.height=6}
sig_smooths <- as.data.frame(summary(gam_season)$s.table) %>%
  mutate(term = rownames(.)) %>%
  filter(grepl("s\\(julian_day\\):station_id", term), `p-value` < 0.05)

sig_sites <- gsub("s\\(julian_day\\):station_id", "", sig_smooths$term)

if (length(sig_sites) > 0) {
  all_pred_grid <- map_dfr(sig_sites, function(site) {
    pred_grid <- tibble(
      julian_day = seq(120, 305, by = 1),
      station_id = factor(site, levels = levels(daily_df$station_id))
    )
    preds <- predict(gam_season, newdata = pred_grid, type = "link", se.fit = TRUE)
    pred_grid %>%
      mutate(
        fit = plogis(preds$fit),
        lower = plogis(preds$fit - 1.96 * preds$se.fit),
        upper = plogis(preds$fit + 1.96 * preds$se.fit),
        station_label = str_remove(site, "^BARKLEYSOUND")
      )
  })

  p_faceted <- ggplot(all_pred_grid, aes(x = julian_day, y = fit)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#A8D5BA", alpha = 0.5) +
    geom_line(color = "#1b9e77", linewidth = 1) +
    labs(x = "Julian Day", y = "Probability of Foraging") +
    facet_wrap(~station_label, ncol = 2) +
    theme_classic(base_size = 14) +
    theme(
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      panel.grid.minor = element_blank()
    )

  print(p_faceted)
  ggsave(here("figures", "gam_seasonal_faceted.png"), plot = p_faceted,
         width = 10, height = 5, dpi = 300)
} else {
  message("No significant site-specific Julian day smooths to plot.")
}
```

# Part B: Tide GAM (bear presence)

## Load tide-joined detections
```{r tide-data}
detections_with_tide <- read_rds(here("data", "processed_data", "detections_with_tide.rds"))

dat0 <- detections_with_tide %>%
  filter(!is.na(tide_height), !is.na(julian_day), !is.na(station_id)) %>%
  mutate(
    bear_presence = as.integer(bear_detected > 0),
    julian_day    = as.numeric(julian_day),
    station_id    = factor(station_id)
  )
```

## Fit tide GAM
```{r tide-gam}
gam_tide <- gam(
  bear_presence ~ s(tide_height, k = 5) + s(julian_day, k = 6) + s(station_id, bs = "re"),
  family = binomial,
  data = dat0,
  method = "REML"
)
summary(gam_tide)

# Reduced model without tide
gam_no_tide <- gam(
  bear_presence ~ s(julian_day, k = 6) + s(station_id, bs = "re"),
  family = binomial,
  data = dat0,
  method = "REML"
)

anova(gam_no_tide, gam_tide, test = "Chisq")
AIC(gam_no_tide, gam_tide)
```

## Diagnostics
```{r tide-diagnostics}
gam.check(gam_tide)
res_gam <- simulateResiduals(gam_tide, n = 1000)
plot(res_gam)
testDispersion(res_gam)
testZeroInflation(res_gam)
```

## Effect plots
```{r tide-plots, fig.width=7, fig.height=4}
tide_seq <- seq(min(dat0$tide_height, na.rm = TRUE), max(dat0$tide_height, na.rm = TRUE), length.out = 200)
season_seq <- seq(min(dat0$julian_day, na.rm = TRUE), max(dat0$julian_day, na.rm = TRUE), length.out = 200)

newdat_tide <- tibble(
  tide_height = tide_seq,
  julian_day  = median(dat0$julian_day, na.rm = TRUE),
  station_id  = levels(dat0$station_id)[1]
)
newdat_season <- tibble(
  tide_height = median(dat0$tide_height, na.rm = TRUE),
  julian_day  = season_seq,
  station_id  = levels(dat0$station_id)[1]
)

pred_tide <- predict(gam_tide, newdata = newdat_tide, type = "link", se.fit = TRUE)
pred_season <- predict(gam_tide, newdata = newdat_season, type = "link", se.fit = TRUE)

plotdat_tide <- tibble(
  x = tide_seq,
  fit = plogis(pred_tide$fit),
  lower = plogis(pred_tide$fit - 1.96 * pred_tide$se.fit),
  upper = plogis(pred_tide$fit + 1.96 * pred_tide$se.fit)
)

plotdat_season <- tibble(
  x = season_seq,
  fit = plogis(pred_season$fit),
  lower = plogis(pred_season$fit - 1.96 * pred_season$se.fit),
  upper = plogis(pred_season$fit + 1.96 * pred_season$se.fit)
)

p_tide <- ggplot(plotdat_tide, aes(x = x, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#1f78b4", alpha = 0.18, color = NA) +
  geom_line(color = "#1f78b4", linewidth = 1) +
  geom_rug(data = dat0, aes(x = tide_height), inherit.aes = FALSE, sides = "b",
           length = unit(0.02, "npc"), alpha = 0.6, colour = "black") +
  labs(x = "Tide height (m)", y = "P(Bear presence)") +
  theme_classic(base_size = 13) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 1))

p_season <- ggplot(plotdat_season, aes(x = x, y = fit)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#e31a1c", alpha = 0.18, color = NA) +
  geom_line(color = "#e31a1c", linewidth = 1) +
  geom_rug(data = dat0, aes(x = julian_day), inherit.aes = FALSE, sides = "b",
           length = unit(0.02, "npc"), alpha = 0.6, colour = "black") +
  labs(x = "Julian day", y = "P(Bear presence)") +
  theme_classic(base_size = 13) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 1))

p_tide
p_season

ggsave(here("figures", "gam_effect_tide_height.png"), plot = p_tide, width = 7, height = 5, dpi = 300)
ggsave(here("figures", "gam_effect_season.png"), plot = p_season, width = 7, height = 5, dpi = 300)
```

# End of Script
